A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-11-07 appendix.tex
dusty epicycles are not real, move to appendix
home/work sync

\section{One-fluid dispersion relation for the streaming instability}\label{compressible_streaming}
We consider an unstratified disk with $\Phi = \Phi (r)$ so that $\p_zP = \p_z\rho =
0$. The background $\p_rP/\rho$, $\tepsilon$ are constant
input parameters. Then for the Epstein drag law, Eq. \ref{epstein}, we
have $\mathcal{C}=0$ in the background state. 
We Fourier analyze in $r$ and $z$ so that $\p_z\to \ii k_z$ and
$\p_r\to \ii k_x$ when acting on perturbations, and denote $|\bm{k}|^2
\equiv k_x^2 + k_z^2$. We consider large $k_x$ and 
%background gradients when compared with that of 
%perturbations. 
neglect the radial density gradient compared to that of pertuburbations 
(formally setting $\p_r\rho=0$). This is also done in most local studies of
dusty disks \citep[e.g.][]{youdin07b}.  
The linearized
equations are, after eliminating the azimuthal velocity:

@@

\subsection{Spuriously growing epicycles}

@@

These growing epicycles have $|\omega|\gtrsim \kappa$. 
However, these modes are absent in full two-fluid
models \citep{youdin05a}. The discrepency lies in the fact that the
first-order  one-fluid equations are only valid for low-frequency waves with
$|\omega|\lesssim \Omega$. Thus only 
low-frequency modes should be retained from analyses based on the
first-order one-fluid approximation.

>>>>

\section{One-fluid dispersion relation for the streaming instability}\label{compressible_streaming}
We consider an unstratified disk with $\Phi = \Phi (r)$ so that $\p_zP = \p_z\rho =
0$. The background $\p_rP/\rho$, $\tepsilon$ are constant
input parameters. Then for the Epstein drag law, Eq. \ref{epstein}, we
have $\mathcal{C}=0$ in the background state. 
We Fourier analyze in $r$ and $z$ so that $\p_z\to \ii k_z$ and
$\p_r\to \ii k_x$ when acting on perturbations, and denote $|\bm{k}|^2
\equiv k_x^2 + k_z^2$. We consider large $k_x$ and background
gradients when compared with that of  perturbations. 
%neglect the radial density gradient compared to that of pertuburbations 
%(formally setting $\p_r\rho=0$). 
 This is also done in most local studies of
dusty disks \citep[e.g.][]{youdin07b}.  
The linearized
equations are, after eliminating the azimuthal velocity:

@@

\subsection{Spuriously growing epicycles}\label{spurious_epi}

@@

These growing epicycles have $|\omega|\gtrsim \kappa$. 
However, these modes are absent in full two-fluid
models \citep{youdin05a}. The discrepency lies in the fact that the
first-order  one-fluid equations are only valid for low-frequency waves with
$|\sigma|\lesssim \Omega$. Thus only 
low-frequency modes should be retained from analyses based on the
first-order one-fluid approximation.

### 2016-11-07 summary.tex
home/work sync

\subsection{Generalization to locally polytropic disks}
We can extend the dusty/adiabatic gas correspondence to 
other fixed equations of state. As an example, consider the locally
polytropic disk 
\begin{align}
  P = K(r,z)\rhog^{\Gamma}, 
\end{align}
where $K$ is a prescribed function and $\Gamma$ is the constant
adiabatic index. Then eliminating $\tepsilon$ from the dust equation
\ref{dusteq} gives 
\begin{align}\label{poly_energy}
  \frac{D P}{D t} &= - \Gamma P\nabla\cdot\bm{v}  + P \bm{v}\cdot\nabla
  \ln{K} + \mathcal{C},\\
  \mathcal{C}& = \frac{\Gamma P}{\rhog}\nabla\cdot\left(\tepsilon\tstop\nabla
  P\right).
\end{align}
Thus the dusty gas behaves like a pure gas with adiabatic index
$\Gamma$. The entropy is given by 
\begin{align}
  \seff = \ln{\left[K^{1/\Gamma}(r,z)\left(1 - \tepsilon\right)\right]}.  
\end{align}
The results of \S\ref{iso_perfect}---\ref{dust_work} remain valid for the strictly 
polytropic disk with $K=$constant, while one sets $c_s^2\to K$ in
\S\ref{dusty_vsi_int}. 
%When $K$ is constant, the corresponding Solberg-Hoiland criteria for
%axisymmetric stability may be obtained from that in standard adiabatic 
%hydrodynamics \citep[e.g.][]{tassoul78} by
%using this definition of entropy in
%Eqs. \ref{dusty_solberg1}---\ref{dusty_solberg2}.

>>>>

\subsection{Generalization to locally polytropic disks}

%\subsubsection{Locally polytropic disks}
We can extend the dusty/adiabatic gas correspondence to 
other fixed equations of state. As an example, consider the locally
polytropic disk 
\begin{align}
  P = K(r,z)\rhog^{\Gamma}, 
\end{align}
where $K$ is a prescribed function and $\Gamma$ is the constant
adiabatic index. Then eliminating $\tepsilon$ from the dust equation
\ref{dusteq} gives 
\begin{align}\label{poly_energy}
  \frac{D P}{D t} &= - \Gamma P\nabla\cdot\bm{v}  + P \bm{v}\cdot\nabla
  \ln{K} + \mathcal{C},\\
  \mathcal{C}& = \frac{\Gamma P}{\rhog}\nabla\cdot\left(\tepsilon\tstop\nabla
  P\right).
\end{align}
Thus the dusty gas behaves like a pure gas with adiabatic index
$\Gamma$. The entropy is given by 
\begin{align}
  \seff = \ln{\left[K^{1/\Gamma}(r,z)\left(1 - \tepsilon\right)\right]}.  
\end{align}
The results of \S\ref{iso_perfect}---\ref{dust_work} remain valid for the strictly 
polytropic disk with $K=$constant, while one sets $c_s^2\to K$ in
\S\ref{dusty_vsi_int}. 
%When $K$ is constant, the corresponding Solberg-Hoiland criteria for
%axisymmetric stability may be obtained from that in standard adiabatic 
%hydrodynamics \citep[e.g.][]{tassoul78} by
%using this definition of entropy in
%Eqs. \ref{dusty_solberg1}---\ref{dusty_solberg2}.

\subsection{Higher order corrections}
{\bf notes for self. maybe exclude in final. }

As discussed in \S\ref{spurious_epi}, the first-order one-fluid 
formulation (Eqs. \ref{masseq}---\ref{tempeq}) can produce
spuriously-growing modes with frequency $|\sigma|\gtrsim \Omega$. This
results from neglecting terms of $O(|\Delta\bm{v}|^2)$. The full set
of one-fluid equations is significantly more complex than our
model \cite[e.g.][]{laibe14}. However, it is possible to improve upon
the first-order equations whilst retaining similarilty to
standard hydrodynamics.

The full one-fluid equations of motion includes the acceleration term 
\begin{align}
%\left.\frac{D\bm{v}}{Dt}\right|_\mathrm{add} = 
\bm{a} =  
-\frac{1}{\rho}\nabla\cdot\left(\frac{\rhod\rhog}{\rho}\Delta\bm{v}\otimes\Delta\bm{v}\right) 
\equiv  \frac{1}{\rho}\nabla\cdot \bm{T},
\end{align}
where $\bm{T}$ is an effective stress tensor. %This term is neglected
%when working to $O(\left|\Delta\bm{v}\right|)$, as we have done
%thusfar.  % In 
% general $\Delta\bm{v}$ obeys an evolutionary equation. In that case, the stress 
% tensor $\bm{T}$ is reminenscent of magnetic tension in
% magneto-hydrodynamics. 
In the terminal velocity approximation with $\Delta\bm{v} = \tstop
\nabla P/\rhog$, we have 
\begin{align}
\bm{T} = - \frac{\epsilon\tstop^2}{\rho}\nabla P \otimes \nabla P.
\end{align}
If we define 
\begin{align}
  \bm{g} \equiv \sqrt{4\pi G \epsilon \rho} \frac{\nabla P}{\rho} \tstop,
\end{align}
then
\begin{align}
\bm{T} = - \frac{1}{4\pi G} \bm{g}\otimes\bm{g}. \label{effective_g}
\end{align}

{\bf tensor virial theorems?}

However, Eq. \ref{effective_g} is just the 
gravitational stress tensor in self-gravitating hydrodynamics
%\footnote{Technically there is an isotropic part, which} 
\citep{lynden-bell72}.

 It is
